# Mammary gland dataset

```{r, echo=FALSE}
knitr::opts_chunk$set(cache=TRUE, error=FALSE, warning=FALSE, message=FALSE)
```

## Introduction

This performs an analysis of the @bach2017differentiation 10X Genomics dataset,
from which we will consider a single sample of epithelial cells from the mouse mammary gland during gestation.

## Analysis code

### Data loading

```{r loading}
library(BiocFileCache)
bfc <- BiocFileCache()
base.path <- "ftp://ftp.ncbi.nlm.nih.gov/geo/samples/GSM2834nnn/GSM2834500/suppl"
barcode.fname <- bfcrpath(bfc, file.path(base.path, 
    "GSM2834500%5FG%5F1%5Fbarcodes%2Etsv%2Egz"))
gene.fname <- bfcrpath(bfc, file.path(base.path,
    "GSM2834500%5FG%5F1%5Fgenes%2Etsv%2Egz"))
counts.fname <- bfcrpath(bfc, file.path(base.path,
    "GSM2834500%5FG%5F1%5Fmatrix%2Emtx%2Egz"))

library(Matrix)
library(SingleCellExperiment)
gene.info <- read.table(gene.fname, stringsAsFactors=FALSE)
colnames(gene.info) <- c("Ensembl", "Symbol")
sce.mam <- SingleCellExperiment(
    list(counts=as(readMM(counts.fname), "dgCMatrix")), 
    rowData=gene.info, 
    colData=DataFrame(Barcode=readLines(barcode.fname))
)
```

### Gene annotation

```{r gene-annotation}
library(scater)
rownames(sce.mam) <- uniquifyFeatureNames(
    rowData(sce.mam)$Ensembl, rowData(sce.mam)$Symbol)

library(AnnotationHub)
ens.mm.v97 <- AnnotationHub()[["AH73905"]]
rowData(sce.mam)$SEQNAME <- mapIds(ens.mm.v97, keys=rowData(sce.mam)$Ensembl,
    keytype="GENEID", column="SEQNAME")
```

### Quality control

```{r quality-control}
is.mito <- rowData(sce.mam)$SEQNAME == "MT"
df <- perCellQCMetrics(sce.mam, subsets=list(Mito=which(is.mito)))
discarded <- quickCellQC(df, percent_subsets="subsets_Mito_percent", nmads=3)
sce.mam <- sce.mam[,!discarded$discard]
```

### Normalization 

```{r normalization}
library(scran)
set.seed(101000110)
clusters <- quickCluster(sce.mam)
sce.mam <- computeSumFactors(sce.mam, clusters=clusters, min.mean=0.1) 
sce.mam <- logNormCounts(sce.mam)
```

### Variance modelling

We use a Poisson-based technical trend to capture more genuine biological variation in the biological component.

```{r variance-modelling}
set.seed(00010101)
dec.mam <- modelGeneVarByPoisson(sce.mam)
```

### Dimensionality reduction

```{r dimensionality-reduction}
library(BiocSingular)
set.seed(101010011)
sce.mam <- denoisePCA(sce.mam, technical=dec.mam, BSPARAM=IrlbaParam())

sce.mam <- runTSNE(sce.mam, dimred="PCA")
```

### Clustering 

We use a higher `k` to obtain coarser clusters (for use in `doubletCluster()` later).

```{r clustering}
snn.gr <- buildSNNGraph(sce.mam, use.dimred="PCA", k=25)
sce.mam$cluster <- factor(igraph::cluster_walktrap(snn.gr)$membership)
```
