# Human pancreas dataset (Grun)

```{r setup, echo=FALSE, results="asis"}
library(OSCAUtils)
chapterPreamble(use_cache = TRUE)
```

## Introduction

This performs an analysis of the @grun2016denovo CEL-seq2 dataset,
consisting of human pancreas cells from various donors.

## Analysis code

### Data loading

```{r loading}
library(scRNAseq)
sce.grun <- GrunPancreasData()
```

### Gene annotation

We convert to Ensembl identifiers, and we remove duplicated genes or genes without Ensembl IDs.

```{r gene-annotation}
library(org.Hs.eg.db)
gene.ids <- mapIds(org.Hs.eg.db, keys=rowData(sce.grun)$symbol,
    keytype="SYMBOL", column="ENSEMBL")

keep <- !is.na(gene.ids) & !duplicated(gene.ids)
sce.grun <- sce.grun[keep,]
rownames(sce.grun) <- gene.ids[keep]
```

### Quality control

This dataset lacks mitochondrial genes so we will do without.

```{r}
unfiltered <- sce.grun
```

```{r quality-control}
library(scater)
stats <- perCellQCMetrics(sce.grun)
qc <- quickPerCellQC(stats, percent_subsets="altexps_ERCC_percent", nmads=3)
sce.grun <- sce.grun[,!qc$discard]
```

### Normalization 

```{r normalization}
library(scran)
set.seed(1000) # for irlba. 
clusters <- quickCluster(sce.grun)
sce.grun <- computeSumFactors(sce.grun, clusters=clusters)
sce.grun <- logNormCounts(sce.grun)
```

### Variance modelling

We block on a combined plate and donor factor.

```{r variance-modelling}
block <- paste0(sce.grun$sample, "_", sce.grun$donor)
dec.grun <- modelGeneVarWithSpikes(sce.grun, spikes="ERCC", block=block)
top.grun <- getTopHVGs(dec.grun, prop=0.1)
```

### Data integration

```{r data-integration}
library(batchelor)
set.seed(1001010)
merged.grun <- fastMNN(sce.grun, subset.row=top.grun, batch=sce.grun$donor)
```

### Dimensionality reduction

```{r dimensionality-reduction}
set.seed(100111)
merged.grun <- runTSNE(merged.grun, dimred="corrected")
```

### Clustering

```{r clustering}
snn.gr <- buildSNNGraph(merged.grun, use.dimred="corrected")
merged.grun$cluster <- factor(igraph::cluster_walktrap(snn.gr)$membership)
```

## Results

### Quality control statistics

```{r, fig.wide=TRUE}
colData(unfiltered) <- cbind(colData(unfiltered), stats)
unfiltered$discard <- qc$discard

gridExtra::grid.arrange(
    plotColData(unfiltered, x="donor", y="sum", colour_by="discard") +
        scale_y_log10() + ggtitle("Total count"),
    plotColData(unfiltered, x="donor", y="detected", colour_by="discard") +
        scale_y_log10() + ggtitle("Detected features"),
    plotColData(unfiltered, x="donor", y="altexps_ERCC_percent",
        colour_by="discard") + ggtitle("ERCC percent"),
    ncol=2
)
```

```{r}
colSums(as.matrix(qc))
```

### Normalization

```{r}
summary(sizeFactors(sce.grun))
```

```{r}
plot(librarySizeFactors(sce.grun), sizeFactors(sce.grun), pch=16,
    xlab="Library size factors", ylab="Deconvolution factors", log="xy")
```

### Variance modelling

```{r, fig.asp=2, fig.height=10}
par(mfrow=c(6,3))
blocked.stats <- dec.grun$per.block
for (i in colnames(blocked.stats)) {
    current <- blocked.stats[[i]]
    plot(current$mean, current$total, main=i, pch=16, cex=0.5,
        xlab="Mean of log-expression", ylab="Variance of log-expression")
    curfit <- metadata(current)
    points(curfit$mean, curfit$var, col="red", pch=16)
    curve(curfit$trend(x), col='dodgerblue', add=TRUE, lwd=2)
}
```

### Data integration

```{r}
metadata(merged.grun)$merge.info$lost.var
```

### Clustering

```{r}
table(Cluster=merged.grun$cluster, Donor=merged.grun$batch)
```

```{r}
plotTSNE(merged.grun, colour_by="cluster")
plotTSNE(merged.grun, colour_by="batch")
```

## Session Info {-}

```{r sessionInfo, echo=FALSE, results='asis'}
prettySessionInfo()
```
