# Human pancreas dataset (Lawlor)

```{r setup, echo=FALSE, results="asis"}
library(OSCAUtils)
chapterPreamble(use_cache = TRUE)
```

## Introduction

This performs an analysis of the @lawlor2017singlecell dataset,
consisting of human pancreas cells from various donors.

## Analysis code

### Data loading

```{r loading}
library(scRNAseq)
sce.lawlor <- LawlorPancreasData()
```

### Gene annotation

```{r gene-annotation}
library(AnnotationHub)
edb <- AnnotationHub()[["AH73881"]]
anno <- select(edb, keys=rownames(sce.lawlor), keytype="GENEID", 
    columns=c("SYMBOL", "SEQNAME"))
rowData(sce.lawlor) <- anno[match(rownames(sce.lawlor), anno[,1]),-1]
```

### Quality control

```{r}
unfiltered <- sce.lawlor
```

```{r quality-control}
library(scater)
stats <- perCellQCMetrics(sce.lawlor, 
    subsets=list(Mito=which(rowData(sce.lawlor)$SEQNAME=="MT")))
qc <- quickPerCellQC(stats, percent_subsets="subsets_Mito_percent")
sce.lawlor <- sce.lawlor[,!qc$discard]
```

### Normalization

```{r normalization}
library(scran)
set.seed(1000)
clusters <- quickCluster(sce.lawlor)
sce.lawlor <- computeSumFactors(sce.lawlor, clusters=clusters)
sce.lawlor <- logNormCounts(sce.lawlor)
```

### Variance modelling

Using age as a proxy for the donor.

```{r variance-modelling}
dec.lawlor <- modelGeneVar(sce.lawlor, block=sce.lawlor$`islet unos id`)
chosen.genes <- getTopHVGs(dec.lawlor, n=2000)
```

### Dimensionality reduction

```{r dimensionality-reduction}
library(BiocSingular)
set.seed(101011001)
sce.lawlor <- runPCA(sce.lawlor, subset_row=chosen.genes, ncomponents=25)
sce.lawlor <- runTSNE(sce.lawlor, dimred="PCA")
```

### Clustering

```{r clustering}
snn.gr <- buildSNNGraph(sce.lawlor, use.dimred="PCA")
sce.lawlor$cluster <- factor(igraph::cluster_walktrap(snn.gr)$membership)
```

## Results

### Quality control statistics

```{r, fig.wide=TRUE}
colData(unfiltered) <- cbind(colData(unfiltered), stats)
unfiltered$discard <- qc$discard

gridExtra::grid.arrange(
    plotColData(unfiltered, x="age", y="sum", colour_by="discard") +
        scale_y_log10() + ggtitle("Total count"),
    plotColData(unfiltered, x="age", y="detected", colour_by="discard") +
        scale_y_log10() + ggtitle("Detected features"),
    plotColData(unfiltered, x="age", y="subsets_Mito_percent",
        colour_by="discard") + ggtitle("Mito percent"),
    ncol=2
)
```

```{r}
plotColData(unfiltered, x="sum", y="subsets_Mito_percent",
    colour_by="discard") + scale_x_log10()
```

```{r}
colSums(as.matrix(qc))
```

### Normalization

```{r}
summary(sizeFactors(sce.lawlor))
```

```{r}
plot(librarySizeFactors(sce.lawlor), sizeFactors(sce.lawlor), pch=16,
    xlab="Library size factors", ylab="Deconvolution factors", log="xy")
```

### Variance modelling

```{r, fig.asp=2, fig.height=10}
par(mfrow=c(4,2))
blocked.stats <- dec.lawlor$per.block
for (i in colnames(blocked.stats)) {
    current <- blocked.stats[[i]]
    plot(current$mean, current$total, main=i, pch=16, cex=0.5,
        xlab="Mean of log-expression", ylab="Variance of log-expression")
    curfit <- metadata(current)
    curve(curfit$trend(x), col='dodgerblue', add=TRUE, lwd=2)
}
```

### Clustering

```{r}
table(sce.lawlor$cluster, sce.lawlor$`cell type`)
```

```{r}
table(sce.lawlor$cluster, sce.lawlor$`islet unos id`)
```

```{r}
plotTSNE(sce.lawlor, colour_by="cluster")
plotTSNE(sce.lawlor, colour_by="islet unos id")
```

## Session Info {-}

```{r sessionInfo, echo=FALSE, results='asis'}
prettySessionInfo()
```
