# Human pancreas dataset (Muraro)

```{r, echo=FALSE}
knitr::opts_chunk$set(cache=TRUE, error=FALSE, warning=FALSE, message=FALSE)
```

## Introduction

This performs an analysis of the @muraro2016singlecell CEL-seq dataset,
consisting of human pancreas cells from various donors.

## Analysis code

### Data loading

```{r loading}
library(scRNAseq)
sce.muraro <- MuraroPancreasData()
```

### Gene annotation

Converting back to Ensembl identifiers.

```{r gene-annotation}
library(AnnotationHub)
edb <- AnnotationHub()[["AH73881"]]
gene.symb <- sub("__chr.*$", "", rownames(sce.muraro))
gene.ids <- mapIds(edb, keys=gene.symb, 
    keytype="SYMBOL", column="GENEID")

# Removing duplicated genes or genes without Ensembl IDs.
keep <- !is.na(gene.ids) & !duplicated(gene.ids)
sce.muraro <- sce.muraro[keep,]
rownames(sce.muraro) <- gene.ids[keep]
```

### Quality control

This dataset lacks mitochondrial genes so we will do without.

```{r}
unfiltered <- sce.muraro
```

```{r quality-control}
library(scater)
stats <- perCellQCMetrics(sce.muraro)
qc <- quickCellQC(stats, nmads=3, percent_subsets="altexps_ERCC_percent")
sce.muraro <- sce.muraro[,!filtered$discard]
```

### Normalization

```{r normalization}
library(scran)
set.seed(1000)
clusters <- quickCluster(sce.muraro)
sce.muraro <- computeSumFactors(sce.muraro, min.mean=0.1, clusters=clusters)
sce.muraro <- logNormCounts(sce.muraro)
```

### Variance modelling

```{r variance-modelling}
# Blocking on a combined plate and donor factor.
block <- paste0(sce.muraro$plate, "_", sce.muraro$donor)
dec.muraro <- modelGeneVarWithSpikes(sce.muraro, "ERCC", block=block)
```

### Dimensionality reduction

```{r dimensionality-reduction}
library(BiocSingular)
set.seed(101011001)
sce.muraro <- denoisePCA(sce.muraro, technical=dec.muraro, BSPARAM=IrlbaParam())
sce.muraro <- runTSNE(sce.muraro, dimred="PCA")
```

### Clustering

```{r clustering}
snn.gr <- buildSNNGraph(sce.muraro, use.dimred="PCA")
sce.muraro$cluster <- factor(igraph::cluster_walktrap(snn.gr)$membership)
```

## Results

### Quality control statistics

```{r, fig.wide=TRUE}
colData(unfiltered) <- cbind(colData(unfiltered), stats)
unfiltered$discard <- qc$discard

gridExtra::grid.arrange(
    plotColData(unfiltered, x="donor", y="sum", colour_by="discard") +
        scale_y_log10() + ggtitle("Total count"),
    plotColData(unfiltered, x="donor", y="detected", colour_by="discard") +
        scale_y_log10() + ggtitle("Detected features"),
    plotColData(unfiltered, x="donor", y="altexps_ERCC_percent",
        colour_by="discard") + ggtitle("ERCC percent"),
    ncol=2
)
```

```{r}
colSums(as.matrix(qc))
```

### Normalization

```{r}
summary(sizeFactors(sce.muraro))
```

```{r}
plot(librarySizeFactors(sce.muraro), sizeFactors(sce.muraro), pch=16,
    xlab="Library size factors", ylab="Deconvolution factors", log="xy")
```

### Variance modelling

```{r, fig.asp=2, fig.height=10}
par(mfrow=c(8,4))
blocked.stats <- dec.muraro$per.block
for (i in colnames(blocked.stats)) {
    current <- blocked.stats[[i]]
    plot(current$mean, current$total, main=i, pch=16, cex=0.5,
        xlab="Mean of log-expression", ylab="Variance of log-expression")
    curfit <- metadata(current)
    points(curfit$mean, curfit$var, col="red", pch=16)
    curve(curfit$trend(x), col='dodgerblue', add=TRUE, lwd=2)
}
```

### Dimensionality reduction

```{r}
ncol(reducedDim(sce.muraro, "PCA"))
```

### Clustering

```{r}
table(sce.muraro$cluster)
```

```{r}
plotTSNE(sce.muraro, colour_by="cluster")
```

## References {-}
