# Chimeric embryo 10X dataset

```{r setup, echo=FALSE, results="asis"}
library(OSCAUtils)
chapterPreamble(use_cache = TRUE)
```

## Introduction

This performs an analysis of the @pijuansala2019single dataset on mouse gastrulation.
Here, we examine chimeric embryos at the E7.5 stage of development 
where td-Tomato-positive embryonic stem cells (ESCs) were injected into a wild-type blastocyst.

## Analysis code

### Data loading

```{r loading}
library(MouseGastrulationData)
sce.chimera <- WTChimeraData(samples=5:10)
sce.chimera
```

```{r, echo=FALSE, results="hide"}
gc()
```

### Feature annotation

```{r feature-annotation}
library(scater)
rownames(sce.chimera) <- uniquifyFeatureNames(
    rowData(sce.chimera)$ENSEMBL, rowData(sce.chimera)$SYMBOL)
```

### Quality control

Quality control on the cells has already been performed by the authors, so we will not repeat it here.
We additionally remove cells that are labelled as stripped nuclei or doublets.

```{r quality-control}
drop <- sce.chimera$celltype.mapped %in% c("stripped", "Doublet")
sce.chimera <- sce.chimera[,!drop]
```

### Normalization

We use the pre-computed size factors in `sce.chimera`.

```{r normalization}
sce.chimera <- logNormCounts(sce.chimera)
```

### Variance modelling

We retain all genes with any positive biological component, to preserve as much signal as possible across a very heterogeneous dataset.

```{r variance-modelling}
library(scran)
dec.chimera <- modelGeneVar(sce.chimera, block=sce.chimera$sample)
chosen.hvgs <- dec.chimera$bio > 0
```

### Merging

We use a hierarchical merge to first merge together replicates with the same genotype, 
and then merge samples across different genotypes.

```{r merging}
library(batchelor)
set.seed(01001001)
merged <- correctExperiments(sce.chimera, 
    batch=sce.chimera$sample, 
    subset.row=chosen.hvgs,
    PARAM=FastMnnParam(
        merge.order=list(
            list(1,3,5), # WT (3 replicates)
            list(2,4,6)  # td-Tomato (3 replicates)
        )
    )
)
```

### Clustering

```{r clustering}
g <- buildSNNGraph(merged, use.dimred="corrected")
clusters <- igraph::cluster_louvain(g)
merged$cluster <- factor(clusters$membership)
```

### Dimensionality reduction

We use an external algorithm to compute nearest neighbors for greater speed.

```{r dimensionality-reduction, cache=TRUE, cache.lazy=TRUE}
merged <- runTSNE(merged, dimred="corrected", external_neighbors=TRUE)
merged <- runUMAP(merged, dimred="corrected", external_neighbors=TRUE)
```

## Results

### Variance modelling

```{r, fig.asp=0.5, fig.wide=TRUE}
par(mfrow=c(1,2))
blocked.stats <- dec.chimera$per.block
for (i in colnames(blocked.stats)) {
    current <- blocked.stats[[i]]
    plot(current$mean, current$total, main=i, pch=16, cex=0.5,
        xlab="Mean of log-expression", ylab="Variance of log-expression")
    curfit <- metadata(current)
    curve(curfit$trend(x), col='dodgerblue', add=TRUE, lwd=2)
}
```

### Batch correction

```{r}
metadata(merged)$merge.info$lost.var
```

### Clustering

```{r}
table(Cluster=merged$cluster, Sample=merged$sample)
```

```{r}
plotTSNE(merged, colour_by="cluster", text_by="cluster", text_col="red")
plotUMAP(merged, colour_by="cluster")
```

## Session Info {-}

```{r sessionInfo, echo=FALSE, results='asis'}
prettySessionInfo()
```
