# PBMC ID 10X dataset (filtered)

```{r, echo=FALSE}
knitr::opts_chunk$set(cache=TRUE, error=FALSE, warning=FALSE, message=FALSE)
```

## Introduction

This performs an analysis of the public PBMC ID dataset generated by 10X Genomics [@zheng2017massively],
starting from the filtered count matrix.

## Analysis code

### Data loading

```{r loading}
library(TENxPBMCData)
pbmcID <- TENxPBMCData('pbmcID')
```

### Quality control

```{r}
unfiltered <- pbmcID
```

```{r quality-control}
is.mito <- grep("MT", rowData(pbmcID)$Symbol_TENx)

library(scater)
stats <- perCellQCMetrics(pbmcID, subsets=list(Mito=is.mito))
high.mito <- isOutlier(stats$subsets_Mito_percent, nmads=3, type="higher")
pbmcID <- pbmcID[,!high.mito]
```

### Normalization

```{r normalization}
pbmcID <- logNormCounts(pbmcID)
```

### Variance modelling

```{r variance-modelling}
library(scran)
decID <- modelGeneVar(pbmcID)
```

```{r feature-selection}
chosen.hvgs <- which(decID$bio > 0)
```

### Dimensionality reduction

```{r dimensionality-reduction}
# Using randomized SVD, which is more efficient for file-backed matrices.
set.seed(10000)
pbmcID <- runPCA(pbmcID, subset_row=chosen.hvgs, ncomponents=25,
    BSPARAM=BiocSingular::RandomParam())

set.seed(100000)
pbmcID <- runTSNE(pbmcID, dimred="PCA")

set.seed(1000000)
pbmcID <- runUMAP(pbmcID, dimred="PCA")
```

### Clustering

```{r clustering}
g <- buildSNNGraph(pbmcID, k=10, use.dimred = 'PCA')
clust <- igraph::cluster_walktrap(g)$membership
pbmcID$cluster <- factor(clust)
```

## Results

### Quality control statistics

```{r, fig.wide=TRUE}
colData(unfiltered) <- cbind(colData(unfiltered), stats)
unfiltered$discard <- high.mito

gridExtra::grid.arrange(
    plotColData(unfiltered, y="sum", colour_by="discard") +
        scale_y_log10() + ggtitle("Total count"),
    plotColData(unfiltered, y="detected", colour_by="discard") +
        scale_y_log10() + ggtitle("Detected features"),
    plotColData(unfiltered, y="subsets_Mito_percent",
        colour_by="discard") + ggtitle("Mito percent"),
    ncol=2
)
```

```{r}
plotColData(unfiltered, x="sum", y="subsets_Mito_percent",
    colour_by="discard") + scale_x_log10()
```

```{r}
summary(high.mito)
```

### Normalization

```{r}
summary(sizeFactors(pbmcID))
```

### Variance modelling

```{r}
plot(decID$mean, decID$total, pch=16, cex=0.5,
    xlab="Mean of log-expression", ylab="Variance of log-expression")
curfit <- metadata(decID)
curve(curfit$trend(x), col='dodgerblue', add=TRUE, lwd=2)
```

### Clustering

```{r}
table(pbmcID$cluster)
```

```{r}
plotTSNE(pbmcID, colour_by="cluster")
```

## References {-}
