# PBMC 10X dataset (unfiltered)

```{r, echo=FALSE}
knitr::opts_chunk$set(cache=TRUE, error=FALSE, warning=FALSE, message=FALSE)
```

## Introduction

This performs an analysis of the public PBMC dataset generated by 10X Genomics [@zheng2017massively],
starting from the unfiltered count matrix.

## Analysis code

### Data loading

```{r loading}
library(BiocFileCache)
bfc <- BiocFileCache("raw_data", ask = FALSE)
raw.path <- bfcrpath(bfc, file.path("http://cf.10xgenomics.com/samples",
    "cell-exp/2.1.0/pbmc4k/pbmc4k_raw_gene_bc_matrices.tar.gz"))
untar(raw.path, exdir=file.path(tempdir(), "pbmc4k"))

library(DropletUtils)
fname <- file.path(tempdir(), "pbmc4k/raw_gene_bc_matrices/GRCh38")
sce.pbmc <- read10xCounts(fname, col.names=TRUE)
```

### Gene annotation

```{r gene-annotation}
library(scater)
rownames(sce.pbmc) <- uniquifyFeatureNames(
    rowData(sce.pbmc)$ID, rowData(sce.pbmc)$Symbol)

library(EnsDb.Hsapiens.v86)
location <- mapIds(EnsDb.Hsapiens.v86, keys=rowData(sce.pbmc)$ID, 
    column="SEQNAME", keytype="GENEID")
```

### Cell detection

```{r cell-detection}
set.seed(100)
e.out <- emptyDrops(counts(sce.pbmc))
sce.pbmc <- sce.pbmc[,which(e.out$FDR <= 0.001)]
```

### Quality control

```{r}
unfiltered <- sce.pbmc
```

```{r quality-control}
stats <- perCellQCMetrics(sce.pbmc, subsets=list(Mito=which(location=="MT")))
high.mito <- isOutlier(stats$subsets_Mito_percent, nmads=3, type="higher")
sce.pbmc <- sce.pbmc[,!high.mito]
```

### Normalization

```{r normalization}
library(scran)
set.seed(1000)
clusters <- quickCluster(sce.pbmc)
sce.pbmc <- computeSumFactors(sce.pbmc, min.mean=0.1, cluster=clusters)
sce.pbmc <- logNormCounts(sce.pbmc)
```

### Variance modelling

```{r variance-modelling}
set.seed(1001)
dec.pbmc <- modelGeneVarByPoisson(sce.pbmc)
```

### Dimensionality reduction

```{r dimensionality-reduction}
set.seed(10000)
sce.pbmc <- denoisePCA(sce.pbmc, technical=dec.pbmc,
    BSPARAM=BiocSingular::IrlbaParam())

set.seed(100000)
sce.pbmc <- runTSNE(sce.pbmc, use_dimred="PCA")

set.seed(1000000)
sce.pbmc <- runUMAP(sce.pbmc, use_dimred="PCA")
```

### Clustering

```{r clustering}
g <- buildSNNGraph(sce.pbmc, k=10, use.dimred = 'PCA')
clust <- igraph::cluster_walktrap(g)$membership
sce.pbmc$cluster <- factor(clust)
```

## Results

### Quality control statistics

```{r, fig.wide=TRUE}
colData(unfiltered) <- cbind(colData(unfiltered), stats)
unfiltered$discard <- high.mito

gridExtra::grid.arrange(
    plotColData(unfiltered, y="sum", colour_by="discard") +
        scale_y_log10() + ggtitle("Total count"),
    plotColData(unfiltered, y="detected", colour_by="discard") +
        scale_y_log10() + ggtitle("Detected features"),
    plotColData(unfiltered, y="subsets_Mito_percent",
        colour_by="discard") + ggtitle("Mito percent"),
    ncol=2
)
```

```{r}
plotColData(unfiltered, x="sum", y="subsets_Mito_percent",
    colour_by="discard") + scale_x_log10()
```

```{r}
summary(high.mito)
```

### Normalization

```{r}
summary(sizeFactors(sce.pbmc))
```

```{r}
plot(librarySizeFactors(sce.pbmc), sizeFactors(sce.pbmc), pch=16,
    xlab="Library size factors", ylab="Deconvolution factors", log="xy")
```

### Variance modelling

```{r}
plot(dec.pbmc$mean, dec.pbmc$total, pch=16, cex=0.5,
    xlab="Mean of log-expression", ylab="Variance of log-expression")
curfit <- metadata(dec.pbmc)
curve(curfit$trend(x), col='dodgerblue', add=TRUE, lwd=2)
```

### Dimensionality reduction

```{r}
ncol(reducedDim(sce.pbmc, "PCA"))
```

### Clustering

```{r}
table(sce.pbmc$cluster)
```

```{r}
plotTSNE(sce.pbmc, colour_by="cluster")
```


