# Mouse brain dataset 

```{r, echo=FALSE}
knitr::opts_chunk$set(cache=TRUE, error=FALSE, warning=FALSE, message=FALSE)
```

## Introduction

This performs an analysis of the @zeisel2015brain C1 dataset,
consisting of mouse brain cells.

## Analysis

### Data loading

We remove repeats and merge together redundant rows for the same gene.

```{r loading}
library(scRNAseq)
sce.zeisel <- ZeiselBrainData()
sce.zeisel <- sce.zeisel[rowData(sce.zeisel)$featureType!="repeat",]

library(scater)
sce.zeisel <- aggregateAcrossFeatures(sce.zeisel, 
    id=sub("_loc[0-9]+$", "", rownames(sce.zeisel)))
```

### Gene annotation

```{r gene-annotation}
library(org.Mm.eg.db)
ensembl <- mapIds(org.Mm.eg.db, keys=rownames(sce.zeisel), 
    keytype="SYMBOL", column="ENSEMBL")
rowData(sce.zeisel)$ENSEMBL <- ensembl
```

### Quality control

```{r}
unfiltered <- sce.zeisel
```

```{r quality-control}
stats <- perCellQCMetrics(sce.zeisel)
qc <- quickCellQC(stats, percent_subsets="altexps_ERCC_percent", nmads=3)
sce.zeisel <- sce.zeisel[,!qc$discard]
```

### Normalization

```{r normalization}
library(scran)
set.seed(1000)
clusters <- quickCluster(sce.zeisel)
sce.zeisel <- computeSumFactors(sce.zeisel, cluster=clusters, min.mean=0.1)
sce.zeisel <- logNormCounts(sce.zeisel)
```

### Variance modelling

```{r variance-modelling}
dec.zeisel <- modelGeneVarWithSpikes(sce.zeisel, "ERCC")
```

### Dimensionality reduction

Note that `denoisePCA()` automatically selects for genes with positive components, 
so explicit specification of `subset.row=` is not required.

```{r dimensionality-reduction}
library(BiocSingular)
set.seed(101011001)
sce.zeisel <- denoisePCA(sce.zeisel, technical=dec.zeisel, BSPARAM=IrlbaParam())
sce.zeisel <- runTSNE(sce.zeisel, dimred="PCA")
```

### Clustering

```{r clustering}
snn.gr <- buildSNNGraph(sce.zeisel, use.dimred="PCA")
sce.zeisel$cluster <- factor(igraph::cluster_walktrap(snn.gr)$membership)
```

## Results

### Quality control statistics

```{r, fig.wide=TRUE}
colData(unfiltered) <- cbind(colData(unfiltered), stats)
unfiltered$discard <- qc$discard

gridExtra::grid.arrange(
    plotColData(unfiltered, y="sum", colour_by="discard") +
        scale_y_log10() + ggtitle("Total count"),
    plotColData(unfiltered, y="detected", colour_by="discard") +
        scale_y_log10() + ggtitle("Detected features"),
    plotColData(unfiltered, y="altexps_ERCC_percent",
        colour_by="discard") + ggtitle("ERCC percent"),
    ncol=2
)
```

```{r}
colSums(as.matrix(qc))
```

### Normalization

```{r}
summary(sizeFactors(sce.zeisel))
```

```{r}
plot(librarySizeFactors(sce.zeisel), sizeFactors(sce.zeisel), pch=16,
    xlab="Library size factors", ylab="Deconvolution factors", log="xy")
```

### Variance modelling

```{r}
plot(dec.zeisel$mean, dec.zeisel$total, pch=16, cex=0.5,
    xlab="Mean of log-expression", ylab="Variance of log-expression")
curfit <- metadata(dec.zeisel)
points(curfit$mean, curfit$var, col="red", pch=16)
curve(curfit$trend(x), col='dodgerblue', add=TRUE, lwd=2)
```

### Dimensionality reduction

```{r}
ncol(reducedDim(sce.zeisel, "PCA"))
```

### Clustering

```{r}
table(sce.zeisel$cluster)
```

```{r}
plotTSNE(sce.zeisel, colour_by="cluster")
```


