---
output:
  html_document
bibliography: ref.bib
---

# Analyzing repertoire sequencing data {#repertoire-seq}

```{r setup, echo=FALSE, results="asis"}
library(rebook)
chapterPreamble()
```

```{r, fig.cap="This page is under construction.", echo=FALSE}
path <- file.path("https://raw.githubusercontent.com/Bioconductor/OSCABase/images/images", "placeholder_yuno.jpg")
knitr::include_graphics(path)
```

## Motivation

An organism's immune repertoire is defined as the set of T and B cell subtypes that contain genetic diversity in the T cell receptor (TCR) components or immunoglobin chains, respectively.
This diversity is important for ensuring that the adaptive immune system can respond effectively to a wide range of antigens.
We can profile the immune repertoire by simply sequencing the relevant transcripts [@georgiou2014promise;@rosati2017overview], a procedure that can be combined with previously mentioned technologies [@zheng2017massively] to achieve single-cell resolution.
This data can then be used to characterize an individual's immune response based on the expansion of T or B cell clones, i.e., multiple cells with the same sequences for each TCR component or immunoglobulin chain.

By itself, single-cell repertoire sequencing data can be readily analyzed with a variety of tools such as those from the [ImmCantation suite](https://immcantation.readthedocs.io/en/stable/).
For example, the `r CRANpkg("alakazam")` package provides functions to perform common analyses to quantify clonal diversity, reconstruct lineages, examine amino acid properties and so on.
We will not attempt to regurgitate their documentation in this chapter; rather, we will focus on how we can integrate repertoire sequencing data structures into our existing `SingleCellExperiment` framework.
This is not entirely trivial as each cell may have zero, one or multiple sequences for any given repertoire component, whereas we only obtain a single expression profile for that cell.

We would like to define a single data structure that captures both the expression profile and repertoire state for each cell.
This ensures synchronization during operations like subsetting (as previously discussed for the `SingleCellExperiment` class) and reduces book-keeping errors throughout the course of an interactive analysis.
We achieve this using the `SplitDataFrameList` class from the `r Biocpkg("IRanges")` package, which allows us to accommodate repertoire sequencing data into existing Bioconductor classes while retaining compatibility with functions from external analysis tools.
We demonstrate on a publicly available 10X Genomics dataset using mouse PBMCs, for which the expression and ADT data have already been processed below:

```{r, echo=FALSE, results="asis"}
extractCached("tenx-repertoire-pbmc8k.Rmd", "clustering", "sce.pbmc")
```

```{r}
sce.pbmc

# Moving ADT-based clustering to the top level for convenience.
colLabels(sce.pbmc) <- colLabels(altExp(sce.pbmc))
```

## Analyzing the T cell receptor repertoire

### Data processing

First, we obtain the filtered TCR contig annotations for the same set of cells.
Each row of the resulting data frame contains information about a single TCR component sequence in one cell, broken down into the alleles of the V(D)J genes making up that component (`v_gene`, `d_gene`, `j_gene`) where possible. 
The number of reads and UMIs supporting the set of allele assignments for a cell is also shown, though only the UMI count should be used for quantifying expression of a particular TCR sequence.
Each cell is assigned to a clonotype (`raw_clonotype_id`) based on the combination of the $\alpha$-chain (TRA) and $\beta$-chain (TRB) sequences in that cell.

```{r}
library(BiocFileCache)
bfc <- BiocFileCache(ask=FALSE)
tcr.data <- bfcrpath(bfc, file.path(
    "http://cf.10xgenomics.com/samples/cell-vdj/3.1.0",
    "vdj_v1_hs_pbmc3/vdj_v1_hs_pbmc3_t_filtered_contig_annotations.csv"))
tcr <- read.csv(tcr.data, stringsAsFactors=FALSE)
nrow(tcr)
head(tcr)
```

The challenge in incorporating all of these data structures into a single object lies in the fact that each cell may have zero, one or many TCR sequences.
This precludes direct storage of repertoire information in the `colData()` of the `SingleCellExperiment`, which would be expecting a 1:1 mapping from each cell to each repertoire sequence.
Instead, we store the repertoire data as a `SplitDataFrameList` object where each cell is represented by a variable-row `DataFrame` containing information for zero-to-many sequences.
This is easily done by:

1. Converting our `data.frame` to a `DataFrame` from the `r Biocpkg("S4Vectors")` package.
We demonstrate this process below for the alpha chain:

    ```{r}
    tra <- tcr[tcr$chain=="TRA",]
    tra <- DataFrame(tra)
    ```

2. Defining a factor of cell identities, using cells that are present in `sce.pbmc` as our levels.
This ensures that, in the resulting `SplitDataFrameList`, we always have one entry per cell in `sce.pbmc`,
even if those entries consist of zero-row `DataFrame`s.

    ```{r}
    cell.id <- factor(tra$barcode, sce.pbmc$Barcode)
    ```

3. Using the `split()` function to break up `tra` into one `DataFrame` per cell, forming the desired `SplitDataFrameList`.

    ```{r}
    tra.list <- split(tra, cell.id)
    class(tra.list)
    ```

Put together, this is as simple as the following stretch of code (repeated for the beta chain):

```{r}
trb <- tcr[tcr$chain=="TRB",]
trb.list <- split(DataFrame(trb), factor(trb$barcode, sce.pbmc$Barcode))
length(trb.list)
```

Both of these objects are guaranteed to have a 1:1 mapping to the columns of `sce.pbmc`, allowing us to directly store them in the `colData` as additional metadata fields.
This ensures that any subsetting applied to the `SingleCellExperiment` is synchronized across both gene expression and repertoire data.

```{r}
sce.pbmc$TRA <- tra.list
sce.pbmc$TRB <- trb.list
```

### Leveraging List semantics

At this point, it is worth spending some time on the power of the `SplitDataFrameList` and the `List` grammar.
In the simplest case, the `SplitDataFrameList` can be treated as an ordinary list of `DataFrame`s with one entry per cell.

```{r}
sce.pbmc$TRA[[1]] # for the first cell.
sce.pbmc$TRA[[2]] # for the second cell.
sce.pbmc$TRA[3:5] # for the third-to-fifth cells.
head(lengths(sce.pbmc$TRA)) # number of sequences per cell.
```

However, it is also possible to treat a `SplitDataFrameList` like a giant `DataFrame` with respect to its columns.
When constructed in the manner described above, all entries of the `SplitDataFrameList` have the same columns;
this allows us to use column-subsetting semantics to extract the same column from all of the internal `DataFrame`s.

```{r}
# Create a new SplitDataFrameList consisting only
# of the columns 'reads' and 'umis'.
sce.pbmc$TRA[,c("reads", "umis")]

# Extract a single column as a new List.
sce.pbmc$TRA[,"reads"]
```

For the `"reads"` column, the above subsetting yields an `IntegerList` where each entry corresponds to the integer vector that would have been extracted from the corresponding entry of the `SplitDataFrameList`.
This is equivalent to looping over the `SplitDataFrameList` and column-subsetting each individual `DataFrame`, though the actual implementation is much more efficient.
The `IntegerList` and its type counterparts (e.g., `CharacterList`, `LogicalList`) are convenient structures as they support a number of vector-like operations in the expected manner.
For example, a boolean operation will convert an `IntegerList` into a `LogicalList`:

```{r}
sce.pbmc$TRA[,"umis"] > 2
```

This is where the final form of `SplitDataFrameList` subsetting comes into play.
If we use a `LogicalList` to subset a `SplitDataFrameList`, we can subset on the individual sequences _within_ each cell.
This is functionally equivalent to looping over both the `LogicalList` and `SplitDataFrameList` simultaneously and using each logical vector to subset the corresponding `DataFrame`. 
For example, we can filter our `SplitDataFrameList` so that each per-cell `DataFrame` only contains sequences with more than 2 UMIs.

```{r}
more.than.2 <- sce.pbmc$TRA[sce.pbmc$TRA[,"umis"] > 2]
head(lengths(more.than.2))
```

We can exploit these semantics to quickly assemble complex queries without the need for any other packages.
Say we want to determine the proportion of cells in each cluster that have at least one productive sequence of a TCR component, i.e., contigs that are likely to produce a functional protein.
Clusters with large counts are most likely to be T cells, though some background level of TCR expression may be observed in other clusters due to a mixture of clustering uncertainty, ambient contamination, doublet formation and potential expression in other cell types.
This quantification is easily achieved with the built-in `List` semantics for both the alpha and beta chains (Figure \@ref(fig:tcr-prop-cluster-prod)).

```{r tcr-prop-cluster-prod, fig.width=10, fig.wide=TRUE, fig.cap="Proportion of cells in each cluster that express at least one productive sequence of the TCR $\\alpha$ (dark) or $\\beta$-chains (light)."}
# Generate a LogicalList class where each entry corresponds to a cell and is
# a logical vector specifying which of that cell's sequences are productive.
is.prod.A <- sce.pbmc$TRA[,"productive"]=="True" 

# We can apply operations to this LogicalList that mimic looping over a logical
# vector. For example, `any()` will return a logical vector of length equal to 
# the number of cells, with a value of TRUE if any sequence is productive.
has.prod.A <- any(is.prod.A)

# And then we simply count the number of cells in each cluster.
tra.counts.prod <- table(colLabels(sce.pbmc)[has.prod.A])

is.prod.B <- sce.pbmc$TRB[,"productive"]=="True"
has.prod.B <- any(is.prod.B)
trb.counts.prod <- table(colLabels(sce.pbmc)[has.prod.B])

ncells <- table(colLabels(sce.pbmc))
barplot(rbind(TRA=tra.counts.prod/ncells, TRB=trb.counts.prod/ncells),
    legend=TRUE, beside=TRUE)
```

Alternatively, we may wish to identify sequences that have UMI counts no less than 50% of the largest UMI count for the same cell.

```{r}
# Assigning to a short-name variable for convenience.
tra <- sce.pbmc$TRA

# Obtaining the maximum UMI count for each cell.
max.umi <- max(tra[,"umis"])

# Generate a LogicalList indicating which sequences have UMI counts
# greater than half the maximum UMI count in the same cell.
keep <- tra[,"umis"] >= max.umi/2

# Examine the distribution of the proportion of sequences in each cell
# that satisfied our requirements.
summary(mean(keep))
```

We can identify sequences that are full-length, productive and have the largest UMI count in the cell.

```{r}
keep <- tra[,"full_length"]=="True" &
    tra[,"productive"]=="True" &
    tra[,"umis"] == max(tra[,"umis"])
tra.sub <- tra[keep]
```

### Converting back to `DataFrame`s

If an operation must be performed on the original sequence-level data frame, 
we can efficiently recover it by calling `unlist()` on our `SplitDataFrameList`.
It is similarly straightforward to regenerate our `SplitDataFrameList` from the data frame by using the `relist()` function.
This framework permits users to quickly switch between sequence level and cell level perspectives of the repertoire data depending on which is most convenient at any given point in the analysis.

```{r}
tra.seq <- unlist(tra)
dim(tra.seq) # Each row represents a sequence now.

# Adding some arbitrary extra annotation (mocked up here).
extra.anno <- DataFrame(anno=sample(LETTERS, nrow(tra.seq), replace=TRUE))
tra.seq <- cbind(tra.seq, extra.anno)

# Regenerating the SplitDataFrameList from the modified DataFrame.
tra2 <- relist(tra.seq, tra)
length(tra2) # Each element represents a cell again.
```

### Further analyses 

## Session Info {-}

```{r sessionInfo, echo=FALSE, results='asis'}
prettySessionInfo()
```
